Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-16 analysis/paper/paper.Rmd
complete discussion and conclusion

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by most metric attributes, such as the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity. In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019]. Among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf. @Roux2015].

- trade [@Alizadeh2018]
- identity, social boundary, visible trait, Raux2015, Boness2017
- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
- shape as an symbol

@@

An increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed. This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions. Northeastern Taiwan had been encountered two major foreign contacts in the 17th century with the Europeans and 19th century with the Chines immigrants. It shows that the standardization of pottery shape appears during the European presence and increased in the Chinese period. This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society. These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes. In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.

>>>>

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by most metric attributes. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity. In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the historical context of Han immigrants [@Wu2019]. At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan. Shape can be viewed as a symbol as an expression the indigenous identity or social boundary because shape is the most obvious and visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time and reached high quantity and diversity in the 19th century, the production of the local pottery continued and became more standardized. This might imply not only the utilitarian function, but an deliberate action to emphasize the pottery tradition as their cultural custom.

@@

This case study demonstrates the application of Elliptic Fourier analysis on ceramic shape to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign culture influences. Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks. Here, EFA approach combined with a significance test for the equality of CVs for shape variables provides a robust way to identify and statistically assess the difference in shape standardization. Lower variation in ceramic shape was identified after the European presence and even much lower variation during the Chinese presence. The direct relationship between foreign influences and highly standardized ceramic shape was tested here on Kiwulan pottery. This finding helps to understand the cause and factor that would lead to the standardization of pottery production in a cross-cultural interaction context. This gave an insight into the discussion of ceramic specialization which is usually associated with the changes in social organization. The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act. The results here suggest an expression of social identity or showing cultural boundary in indigenous societies through visible variable when encountering foreign intrusion. The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context. Our case study includes an openly available pipeline suitable for other assemblages, ensuring reproducibility and expanding the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
